---
title: "Advanced Bioinformatics 2026 assessment"
author: "M2205233"
date: "2026-04-12"
output: html_document
---

#General R/Rstudio assessment 

##Task 3.1. 
 Sum of all integers between 5 and 55
 The : operator creates a sequence of integers from 5 to 55.
 sum() then adds up all numbers in that sequence.


```{r}
sum(5:55)
#expected 1530
```

##Task 3.2. 

```{r}
#Function sumfun(n) to sum integers from 5 to n
sumfun <- function(n) {sum(5:n)}

sumfun(10)      #expected 45
sumfun(20)      #expected 200
sumfun(100)     #expected 5040
```

##Task 3.3 
First 12 Fibonacci numbers using a for loop
```{r}
fib <- numeric(12)        # create numeric vector of length 12
fib[1] <- 1               # first term
fib[2] <- 1               # second term

for (i in 3:12) {
  fib[i] <- fib[i - 1] + fib[i - 2]
}

print(fib)
# Expected: 1 1 2 3 5 8 13 21 34 55 89 144
```

##Task 3.4
 Boxplot of mpg by gear from mtcars using ggplot2

```{r}
library(ggplot2)

# Make gear a factor for a categorical x-axis

mtcars$gear <- factor(mtcars$gear)

ggplot(mtcars, aes(x = gear, y = mpg, fill = gear)) +
  geom_boxplot() +
  labs(
    title = "Miles per Gallon by Number of Gears",
    x = "Number of Gears",
    y = "Miles per Gallon"
  )
```

##Task 3.5
Linear relationship between speed and stopping distance

```{r}
data(cars)   # loads the built-in 'cars' dataset

fit <- lm(dist ~ speed, data = cars)
summary(fit)

#Intercept ≈ −17.58 (SE ≈ 6.76)
#Slope ≈ 3.93 (SE ≈ 0.42)
```
 
##Task 3.6
```{r}
library(ggplot2)

ggplot(cars, aes(x = speed, y = dist)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "purple") +
  labs(
    title = "Stopping Distance vs Speed with Linear Fit",
    x = "Speed (mph)",
    y = "Stopping distance (ft)"
  )
```

##Task 3.7
```{r}
quad_fit <- lm(dist ~ speed + I(speed^2), data = cars)

library(ggplot2)

ggplot(cars, aes(x = speed, y = dist)) +
  geom_point() +  # observed data
  stat_smooth(
    method  = "lm",
    formula = y ~ x + I(x^2),
    se      = FALSE,
    color   = "Purple"
  ) +
  labs(
    title = "Stopping Distance vs Speed with Quadratic Fit",
    x = "Speed (mph)",
    y = "Stopping distance (ft)"
  )

```

#RNA-seq assessment 

##Task3.8
```{r}
# Read the count matrix
all_counts <- read.csv(file = "/Users/aisha/Downloads/LMS_RNAseq_short-master-2023-final-2/course/exercises/data/exercise1_counts.csv", header = T, row.names = 1)
head(all_counts)

# Read the sample description file

sam_des <- read.table("/Users/aisha/Downloads/LMS_RNAseq_short-master-2023-final-2/course/exercises/data/exercise1_sample_description.info", sep = "\t", header = TRUE)
head(sam_des)


```

##Task 3.9

```{r}
# Create col data from sample description
col_data <- data.frame(sample = sam_des$sample, condition = sam_des$condition, batch = sam_des$batch)
head(col_data)

# Check dimensions
all(colnames(all_counts) == col_data$name)
```

##Task 3.10
```{r,collapse=T,warning=FALSE,eval=TRUE}

library(DESeq2)
# Create DESeq object
dds <- DESeqDataSetFromMatrix(countData = all_counts, colData = col_data, design =~ condition)
dds

```

##Task 3.11

```{r}
# rlog transformation
rld <- rlog(dds, blind = TRUE)


# VST transformation
vsd <- varianceStabilizingTransformation(dds, blind = TRUE)
```

##Task 3.12

```{r}

# Load libraries
library("pheatmap")
library(DESeq2)

# Get dds normalized counts
dds <- DESeq(dds)
dds_counts <- counts(dds, normalized = TRUE)
head(dds_counts)

# Select top 40 highly expressed genes
select <- order(rowMeans(counts(dds, normalized = TRUE)), decreasing = TRUE)[1:40]

# Define select FROM the SAME transformation you're plotting
rld_mat <- assay(rld)
select_rld <- order(rowMeans(rld_mat), decreasing = TRUE)[1:40]

vsd_mat <- assay(vsd)
select_vsd <- order(rowMeans(vsd_mat), decreasing = TRUE)[1:40]

# CREATE annotation_col (MUST match your colData)
annotation_col <- data.frame(
  condition = colData(rld)$condition,  # Replace 'condition' with your actual column name
  row.names = colnames(rld)
)

# Heatmap for rlog data
pheatmap(assay(rld)[select, ],
         scale = "row",
         clustering_distance_rows = "correlation",
         clustering_distance_cols = "correlation",
         show_rownames = FALSE,
         annotation_col = annotation_col,
         main = "Top 40 expressed genes - rlog")

# Heatmap for VST data

pheatmap(assay(vsd)[select, ],
         scale = "row",
         clustering_distance_rows = "correlation",
         clustering_distance_cols = "correlation",
         show_rownames = FALSE,
         annotation_col = annotation_col,
         main = "Top 40 highly expressed genes - VST")

         
```


##Task 3.13

```{r}
# Compute SDM from rlog transformed data
sample_dist <- dist(t(assay(rld)))
class(sample_dist)

# Get SDM in matrix form
sdm <- as.matrix(sample_dist)
class(sdm)

# Load library
library("RColorBrewer")

# Add row names for clear plot
rownames(sdm) <- rld$Group
colnames(sdm) <- NULL

# Add colors
colors <- colorRampPalette(rev(brewer.pal(9, "Reds")))(255)

# Plot heatmap
pheatmap(sdm,
         clustering_distance_rows = sample_dist,
         clustering_distance_cols = sample_dist,
         col = colors)
```

##Task 3.14

```{r}
library(DESeq2)

# Generate rlog-transformed data
rld <- rlog(dds, blind = TRUE)

# Perform PCA and extract variance percentages
pcaData <- plotPCA(rld, intgroup = "condition", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
percentVar

#plot
library(ggplot2)

ggplot(pcaData, aes(PC1, PC2, color = condition)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal()
```



##Task 3.15

```{r}
# Generate rlog-transformed data
vsd <- vst(dds, blind = TRUE)

# Perform PCA and extract variance percentages
pcaData_vst <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
percentVar_vst <- round(100 * attr(pcaData_vst, "percentVar"))
percentVar_vst

#save plot
library(ggplot2)
ggplot(pcaData_vst, aes(PC1, PC2, color = condition)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar_vst[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_vst[2], "% variance")) +
  theme_minimal()
```


